Pre-screening comments

### 2016-03-07 manuscript2.tex
more JSS style conformation, expanded examples
added package citations (and bibtex JSS style changes); added data
descriptives like ?bscoup, str(bscoup); changed bibtex titles to title
case; added a few more \pkg’s and \proglang’s, esp. in bibs;
replication runs through on new R session

%% almost as usual
\author{
Andreas Beger\\Duke University \And Daniel W. Hill, Jr.\\University of Georgia \And Nils W. Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting it Up: The \pkg{spduration} Split-population Duration
Regression Package}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
\Plaintitle{Splitting it up: the spduration split-population duration regression
package}
\Shorttitle{\pkg{spduration}: Split-population duration regression}

%% an abstract and keywords
\Abstract{
We present an implementation of split-population duration regression in
the \pkg{spduration} \proglang{R} package and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models.
}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("spduration")
R> data(bscoup)
R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
+    tID = "year", freq = "year", ongoing = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data and will try to convert the
\code{tID} input to class ``Date'' given the provided argument in
\code{freq}. Another important question is how to handle consecutive 1's
in the response variable provided. This is controlled with the
\code{ongoing} argument. In some cases like civil war occurrence, the
\emph{y} variable records all years during which a country experienced a
civil war. Here \code{ongoing} should be set to true so that the models
try to predict the onset of the civil war, but disregard ongoing
conflicts (\code{failure} is set to \code{NA} for these cases, dropping
them from analysis). With ongoing set to false, successive 1's in
\emph{y} are treated as distinct and new failures, and kept in the
analysis. This makes sense for discrete, short-term events like coups.
Countries can experience distinct coups in successive years.

@@

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. These can be
exported by calling \code{xtable} directly on the ``spdur'' object to
produce Table 2.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
  milreg & $-$1.13 & 0.21 & 0.00 \\ 
  instab & $-$0.09 & 0.02 & 0.00 \\ 
  regconf & $-$2.52 & 2.16 & 0.24 \\ 
  log(alpha) & $-$0.45 & 0.06 & 0.00 \\ 
  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
  couprisk & 0.59 & 0.32 & 0.07 \\ 
  wealth & $-$0.36 & 0.28 & 0.20 \\ 
  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
  rwar & $-$0.52 & 0.94 & 0.58 \\ 
  regconf.1 & $-$5.43 & 5.62 & 0.33 \\ 
  samerica & 2.09 & 1.45 & 0.15 \\ 
  camerica & $-$0.39 & 0.73 & 0.59 \\ 
   \hline
\end{tabular}
\caption{Coup model with log-logistic hazard.} \label{loglog_table}
\end{table}

@@

The package also includes a function, \code{plot_hazard}, that can be
used with an \code{spdur} object to plot the estimated hazard rate.
\code{plot_hazard} plots the conditional hazard, which is the
probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\). The
function calculates the estimated hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters. By default the function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \code{xvals} and \code{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates the graph A in Figure \ref{hazard-ex}.

@@

The hazard plots can also be called with
\code{plot(..., type = "hazard")}. Without a type argument,
\code{plot.spdur} will produce a separation plot, which is a graphical
display for evaluating model predictions \citep{greenhill2011separation}. The code below produces Figure \ref{insamp}.

@@

\section{Censoring considerations}

@@

\section{Alternative implementations}

The \pkg{survival} library, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data. It does not include cure rate or split-population mixture models of the type we have implemented though.

>>>>

%% almost as usual
\author{
Andreas Beger\\Duke University \And Daniel W. Hill, Jr.\\University of Georgia \And Nils W. Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting It Up: The \pkg{spduration} Split-Population Duration
Regression Package}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
\Plaintitle{Splitting It Up: The spduration Split-Population Duration Regression
Package}
\Shorttitle{\pkg{spduration}: Split-Population Duration Regression}

%% an abstract and keywords
\Abstract{
We present an implementation of split-population duration regression in
the \pkg{spduration} \citep{beger2016spduration} package for \proglang{R} \citep{rstats2016} and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models.
}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("spduration")
R> data(bscoup)
R> ?bscoup
R> head(bscoup[, 1:5])
\end{CodeInput}
\begin{CodeOutput}
  countryid year  couprisk recentcoups rwar
1       909 1960 0.2200166           0    0
2       909 1961 0.2200166           0    0
3       909 1962 0.2200166           0    0
4       909 1963 0.2200166           0    0
5       909 1964 2.5702426           1    0
6       909 1965 2.5702426           2    0
\end{CodeOutput}
\end{CodeChunk}

The data are documented in `?bscoup`, which also gives a reference to the source citation. It consists of a little more than 5,000 observations of 162 countries from 1960 to 2000. Each row corresponds to a country $c$ in year $t$. Excluding the country and year identifiers, the data include 12 variables, including a binary indicator for successful coups.

\begin{CodeChunk}
\begin{CodeInput}
R> str(bscoup)
\end{CodeInput}
\begin{CodeOutput}
'data.frame':	5463 obs. of  14 variables:
 $ countryid  : num  909 909 909 909 909 909 909 920 920 920 ...
 $ year       : num  1960 1961 1962 1963 1964 ...
 $ couprisk   : num  0.22 0.22 0.22 0.22 2.57 ...
 $ recentcoups: num  0 0 0 0 1 2 4 0 0 0 ...
 $ rwar       : num  0 0 0 0 0 0 1 0 0 0 ...
 $ milreg     : num  0 0 0 1 0 1 1 0 0 0 ...
 $ wealth     : num  NA NA NA NA NA ...
 $ instab     : num  4 2 9 9 12 13 12 NA 0 4 ...
 $ coup       : Factor w/ 2 levels "no","yes": 1 1 1 2 2 2 1 1 1 1 ...
 $ africa     : num  0 0 0 0 0 0 0 0 0 0 ...
 $ eurnam     : num  0 0 0 0 0 0 0 1 1 1 ...
 $ samerica   : num  0 0 0 0 0 0 0 0 0 0 ...
 $ camerica   : num  0 0 0 0 0 0 0 0 0 0 ...
 $ regconf    : num  0 0 0 0.019 0.019 ...
 - attr(*, "datalabel")= chr ""
 - attr(*, "time.stamp")= chr "30 Jul 2003 12:15"
 - attr(*, "formats")= chr  "%9.0g" "%9.0g" "%9.0g" "%9.0g" ...
 - attr(*, "types")= int  102 102 102 102 102 102 102 102 102 102 ...
 - attr(*, "val.labels")= chr  "" "" "" "" ...
 - attr(*, "var.labels")= chr  "country code .pr89" "year coded" "" "" ...
 - attr(*, "version")= int 7
 - attr(*, "label.table")=List of 1
  ..$ hadcoup: Named int  0 1
  .. ..- attr(*, "names")= chr  "no" "yes"
\end{CodeOutput}
\begin{CodeInput}
R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
R> table(bscoup$coup)
\end{CodeInput}
\begin{CodeOutput}
   0    1
5173  290 
\end{CodeOutput}
\end{CodeChunk}

Alltogether the data include 290 coups, which makes for a positive rate of slightly more than 5\% of all observations.

Before we can estimate a split-population duration model with the data, we need to add several variables that capture the survival characteristics of the data.

\begin{CodeChunk}
\begin{CodeInput}
R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
+    tID = "year", freq = "year", ongoing = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning message:
In attempt_date(data[, tID], freq) :
  Converting to 'Date' class with yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

@@

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data and will try to convert the
\code{tID} input to class ``Date'' given the provided argument in
\code{freq}, as indicated by the warning it returned in the example above. This 
can be avoided by converting any numeric dates to \proglang{R} \code{Date}s first.

Another important question is how to handle consecutive 1's
in the response variable provided. This is controlled with the
\code{ongoing} argument. In some cases like civil war occurrence, the
\emph{y} variable records all years during which a country experienced a
civil war. Here \code{ongoing} should be set to true so that the models
try to predict the onset of the civil war, but disregard ongoing
conflicts (\code{failure} is set to \code{NA} for these cases, dropping
them from analysis). With ongoing set to false, successive 1's in
\emph{y} are treated as distinct and new failures, and kept in the
analysis. This makes sense for discrete, short-term events like coups.
Countries can experience distinct coups in successive years.

@@

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. These can be
exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
produce Table 2.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Parameter & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
Dur\_(Intercept) & 2.40 & 0.21 & 11.43 & 0.00 \\ 
  Dur\_milreg & $-$1.13 & 0.21 & $-$5.46 & 0.00 \\ 
  Dur\_instab & $-$0.09 & 0.02 & $-$4.79 & 0.00 \\ 
  Dur\_regconf & $-$2.52 & 2.16 & $-$1.16 & 0.24 \\ 
  log(alpha) & $-$0.45 & 0.06 & $-$7.09 & 0.00 \\ 
  Risk\_(Intercept) & 2.93 & 1.87 & 1.57 & 0.12 \\ 
  Risk\_couprisk & 0.59 & 0.32 & 1.82 & 0.07 \\ 
  Risk\_wealth & $-$0.36 & 0.28 & $-$1.29 & 0.20 \\ 
  Risk\_milreg & 10.82 & 9.31 & 1.16 & 0.25 \\ 
  Risk\_rwar & $-$0.52 & 0.94 & $-$0.56 & 0.58 \\ 
  Risk\_regconf & $-$5.43 & 5.62 & $-$0.97 & 0.33 \\ 
  Risk\_samerica & 2.09 & 1.45 & 1.45 & 0.15 \\ 
  Risk\_camerica & $-$0.39 & 0.73 & $-$0.53 & 0.59 \\ 
   \hline
\end{tabular}
\caption{Coup model with log-logistic hazard.} 
\label{loglog_table}
\end{table}

@@

The package includes two types of plots that show the estimated hazard rates and 
predictive performance, respectively. The hazard rates can be plotted by either 
calling \code{plot_hazard} directly, or with \code{plot(x, type = "hazard")}, on
a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\), when holding
all covariates at their sample means. The
function calculates the average hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters.

\begin{CodeChunk}
\begin{CodeInput}
R> plot(weib_model,   type="hazard", main="Weibull")
R> plot(loglog_model, type="hazard", main="Loglog")
\end{CodeInput}
\end{CodeChunk}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-rates.pdf}
\caption{Hazard rate plots for the Weibull and Loglogistic coup models.}
\label{hazard-ex}
\end{center}
\end{figure}

By default the \code{plot_hazard} function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \code{xvals} and \code{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates the graph A in Figure \ref{hazard-ex}.

@@

While \code{plot(x, type = "hazard")} will produce a hazard rate plot, without a type argument,
\code{plot.spdur} will produce a separation plot. Separation plots are a graphical
display for evaluating model predictions \citep{greenhill2011separation}. The code below produces Figure \ref{insamp}.

@@

\section{Additional discussion}

\subsection{Censoring considerations}

@@

\subsection{Comparison with other software}

The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguable the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variet of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data. It does not include cure rate or split-population mixture models of the type we have implemented though.

### 2016-03-07 pre-screening-comments.Rmd
more JSS style conformation, expanded examples
added package citations (and bibtex JSS style changes); added data
descriptives like ?bscoup, str(bscoup); changed bibtex titles to title
case; added a few more \pkg’s and \proglang’s, esp. in bibs;
replication runs through on new R session

---
title: "Pre-screening comments"
author: "Andreas Beger"
date: "March 6, 2016"
output: pdf_document
---

@@

*Changed `And` to `\AND`, added `R>` and `+` where needed to code formatting, and bibliography is not via BibTeX.*

@@

@@

*Added discussion of `smcure`, which is the only alternative implementation for R that we could find.*

@@

@@

@@

@@

>>>>

---
title: "Pre-screening comments"
author: "The Authors"
date: "March 6, 2016"
output: pdf_document
---

@@

*Changed `And` to `\AND`, added `R>` and `+` where needed to code formatting, and bibliography is now via BibTeX.*

@@

*Changed `-` to `$-$`, and the column names in `xtable.spdur` now match those output by the `xtable` internal methods, e.g. `xtable.lm`.*

@@

*Added discussion of `smcure`, which is the only alternative implementation for R that we could find, as well `survival`, which handles non-mixture duration models.*

@@

*The two included data sets are documented in the package, e.g. `?coups`. We added some code and examples to the paper describing the data in some more details.*

@@

*Done, and the same for all citations.*

@@

*The code examples have spaces before/after operators and after commas.*

@@

*Checked and added where needed, including in the references.*
